Scheduling and Amortization for MPC

### 2020-08-26 paper.tex
Some changes.

% !TeX TXS-program:compile = txs:///pdflatex/[--shell-escape]
%
% acm templates v1.57 (sigconf template)
%
%\documentclass[sigconf, screen, review, anonymous, natbib=false]{acmart}
\documentclass[sigconf, screen, natbib=false, dvipsnames, table]{acmart}
\settopmatter{printacmref=true}
% mandatory for CCS'19
\usepackage{balance}
% for creating a balanced last page (usually last page with references)

@@

\usepackage{booktabs} % better tables

@@

%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title[Scheduling and Parallelization for MPC]{Scheduling and Parallelization for MPC}

@@

@@

\section{Scheduling in MPC}

@@

\section{Scheduling in HPC}

@@

\ana{Here we have to make the argument that MPC is different from HPC.}
The example illustrates that optimal MPC scheduling and optimal HPC scheduling are different problems.

@@

@@

\section{Optimal Loop Parallelization} 
\label{sec:optimal_loop_parallelization}
%\input{../sections/optimal_loop_parallelization.tex}

\ana{Aiken's and Array SSA to define the notion of optimal loop parallelization for MPC.}

@@

\subsection{Problem Overview}

Recall Problem 2, given a loop-body schedule, compute the across-loop schedule
with minimal cost. The following examples illustrate. The first example is the inner 
product computation, and the second is Aiken and Nicolau's example.

\begin{tabular}{lll}
\begin{lstlisting}[language=Java]
A = ...
B = ...
ip = 0;
for (i=1; i<=N; i++) {
  t=A[i]*B[i];
  ip=ip+t;
}

\end{lstlisting}
& 
~
&
\begin{lstlisting}[language=Java]
A = ...
B = ...
ip = 0;
for (i=1; i<=N; i++) {
  if (i%step5==1) para(t[i],...,t[i+step])
  if (i%step6==1) ip = ip + t[i];
}  
\end{lstlisting}

@@

\ana{More to introduce what the schedule is.}

In the above example {\sf step5} is N, meaning that all N multiplications are executed in parallel.
In contrast, {\sf step6} is 1, meaning that the summation statements are executed sequentially.

\begin{tabular}{lll}
\begin{lstlisting}[language=Java]
for (i=1; i<=N; i++) {
  A[i]=f1(B[i]);
  B[i]=f2(A[i],D[i-1]);
  C[i]=f3(A[i],D[i-1]);
  D[i]=f4(B[i],C[i]);
}

\end{lstlisting}
& 
~
&
\begin{lstlisting}[language=Java]
for (i=1; i<=N; i++) {
  if (i%N==1) para(A[i]=f1(B[i]),...);
  if (i%1==1) para(B[i]=f2(A[i],D[i-1]),...);
  if (i%1==1) para(C[i]=f3(A[i],D[i-1]),...);
  if (i%1==1) para(D[i]=f4(B[i],C[i]),...);
}	
\end{lstlisting}

The above across-loop schedule means that the first instruction is parallelized---all
N instructions A can be computed at once because each depends only on {\sf B[i]}, which
is assigned outside of the loop. In contrast, subsequent instructions are executed 
sequentially. Take B, for example. There is a dependence ``cycle'' from {\sf B[i]} 
to {\sf D[i]} then {\sf B[i+1]} in the subsequent iteration. (The cycle is shown in~\figref{fig:MPCexample}(b).) 
{\sf B[i]}'s cannot be computed in parallel because each one depends on the {\sf B[i]} 
value computed in the previous iteration.

Now consider the above code but slightly modified. {\sf B[i]} and {\sf C[i]} are now 
computed based on the value of {\sf D[i]} from two iterations ago.

\end{lstlisting}
& 
~
&
\begin{lstlisting}[language=Java]
for (i=1; i<=N; i++) {
  if (i%N==1) para(A[i]=f1(B[i]));
  if (i%2==1) para(B[i]=f2(A[i],D[i-2]));
  if (i%2==1) para(C[i]=f3(A[i],D[i-2]));
  if (i%2==1) para(D[i]=f4(B[i],C[i]));
}	
\end{lstlisting}

{\sf B[i]} still depends on values computed in earlier iterations, but one of the values, D[i-2] 
is two iterations earlier. Thus, we can take parallelization steps of 2, e.g., computing 
{\sf B[1]} and {\sf B[2]} in parallel, then {\sf B[3]} and {\sf B[4]} in parallel, etc.
Note however, that we cannot take a parallelization step larger than 2.
The value of B will depend on a value being computed in parallel.

To simplify, we assume that a loop-body schedule is a straight line. \ana{What does this mean?}
\ana{Will we relax?}
The problem therefore becomes: compute the maximal parallelization 
step allowed for each statement. This will account for an MPC schedule with
minimal cost. \ana{Why?}

@@

\paragraph{MPC-Source.} MPC-Source is an SSA-based intermediate representation
of MPC programs~\cite{ccs paper}. MPC-Source is, essentially, a straight-line sequence
of ``components'', where each ``component'' is either a nested for-loop sequence (called for-loop block), 
or a simple MPC-statement. \figref{fig:example} shows an example. MPC-Source can be constructed
in a standard way from an SSA-translation of the source---each branch of an 
{\sf if}-statement is added to the sequence, the conditional becomes an MPC comparison (CMP), 
and the $\phi$-node becomes an MPC multiplexer (MUX).

\paragraph{Array SSA.} The other essential concept we build upon is Array SSA. Originally proposed

\ana{3. MPC-Source with Array SSA}

\begin{lstlisting}[language=Java,escapeinside=`']
for (i=1; i<=N; i++) {

@@

  A2[i] = f1(B1[i]); @A2[i] = i;
  A3 = `$\phi$'(A2,A1); @A3 = max(A2, A1);
  B2[i] = f2(A3[i],D1[i-1]); @B2[i] = i;
  B3 = `$\phi$'(B2,B1); @B3 = max(B2, B1);
  C3[i] = f3(A3[i],D1[i-1]); @C2[i] = i;
  C3 = `$\phi$'(C2,C1); @C3 = max(C2, C1);
  D3[i] = f4(B3[i],C3[i]); @D2[i] = i;
  D3 = `$\phi$'(D2,D1); @D3 = max(D2, D1); 
}
\end{lstlisting}

"Index invariant" makes an assertion about when A[i] is written as a function of i.
\ana{Figure out how to reason about cycles based on Array SSA.}

\subsection{Algorithm}

for a statement x = f(y,z), follow the reverse def-use link.

>>>>

% !TeX TXS-program:compile = txs:///pdflatex/[--shell-escape]
%
% acm templates v1.57 (sigconf template)
%
%\documentclass[sigconf, screen, review, anonymous, natbib=false]{acmart}
\documentclass[sigconf, screen, natbib=false, dvipsnames, table]{acmart}
\settopmatter{printacmref=true}
% mandatory for CCS'19
\usepackage{balance}
\usepackage{mathtools}
% for creating a balanced last page (usually last page with references)

@@

\usepackage{booktabs} % better tables
\usepackage{algorithm}
\usepackage{algorithmic}

@@

%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title[Scheduling and Amortization for MPC]{Scheduling and Amortization for MPC}

@@

\renewcommand{\A}{{\sf A}}
\renewcommand{\B}{{\sf B}}
\renewcommand{\C}{{\sf C}}
\renewcommand{\D}{{\sf D}}

@@

\section{Background and Problem Statement}
\label{sec:background}

\subsection{Scheduling in MPC}

\ana{Here we need more on MPC before jumping to scheduling. MPC-source, basic assumptions about static loop bounds, etc.}

@@

\ana{Mention difficulties. We start from IMP-like general-purpose PL syntax with array reads and writes. We translate the program into a dependence 
graph of MPC instructions. Then we proceed to compute the schedule.}

\subsection{Scheduling in HPC}
\label{sec:hpc}

@@

\ana{Here we have to make the argument that MPC is different from HPC.}
The example illustrates that optimal MPC scheduling and optimal HPC scheduling are different problems.
To summarize, scheduling and amortization in MPC is a different problem due to two key differences: (1) there 
is unlimited bandwidth while in HPC, we typically have a limited number of processors we can use to 
schedule instructions in parallel, and (2) we benefit from amortization only if the set of instructions scheduled
in parallel are the same; in contrast, in HPC any two instructions can be scheduled in parallel.

\ana{Have to make the argument that in MPC all stmts are executed. Thus, the vectorization problem is
immediately simpler!}

@@

\subsection{Precise Problem Statement}
\ana{TODO}

@@

\section{Across-Loop Parallelization} 
\label{sec:optimal_loop_parallelization}
%\input{../sections/optimal_loop_parallelization.tex}

@@

\begin{comment}
\subsection{Overview and Problem Statement}

The problem is, given a loop-body schedule, compute the across-loop schedule
with minimal cost. The following examples illustrate. The first example is the inner 
product computation, and the second is Aiken and Nicolau's example.

\begin{center}
\begin{lstlisting}[language=Java]
A = ...
B = ...
ip = 0;
for (i=0; i<N; i++) {
  if (i%d(5)==0) para{t[i]=A[i]*B[i],...,t[i+d(5)-1]=A[i+d(5)-1]*B[i+d(5)-1]}
  if (i%d(6)==0) para{ip = ip + t[i]};
}  
\end{lstlisting}
\end{center}

The above notation is a convenient way to specify a schedule. Each statement $s$ in the loop body sequence 
is parallelized at a rate $d(s)$ (we assume that the loop body is a straight-line sequence which is the case in MPC). 
Notation  
\[{\sf if \; (i\%d(s)\!==\!0) \; para\{s(i),s(i+1),...,s(i+d(s)-1)\}}\]
means that the first $d(s)$ iterations of statement $s$, denoted by $s(i),...s(i+d(s)-1)$, are executed in a parallel block; 
then the following $d(s)$ iterations are executed in a parallel block, and so on. The execution of $d(s)$ statements $s$ in 
a block benefits from amortization of the operation at $s$.
If {\sf i\%d(s)==0} evaluates to false, then the $i$-th iteration skips $s$, because $s(i)$ has executed earlier in a parallel block.

In the above example {\sf d(5)} is N, meaning that all N multiplications are executed in parallel.
In contrast, {\sf d(6)} is 1, meaning that the summation statements are executed sequentially.

\begin{lstlisting}[language=Java]
for (i=0; i<N; i++) {
  if (i%N==0) para{A[i]=f1(B[i]),...,A[i+N-1]=f1(B[i+N-1])};
  if (i%1==0) para{B[i]=f2(A[i],D[i-1])};
  if (i%1==0) para{C[i]=f3(A[i],D[i-1])};
  if (i%1==0) para{D[i]=f4(B[i],C[i])};
}	
\end{lstlisting}

The above across-loop schedule entails that the first instruction is parallelized---all
N instructions A can be computed at once because each depends only on {\sf B[i]}, 
which at the point {\sf A[i]} is computed, refers to a value assigned outside of the loop. 
In contrast, subsequent instructions are executed sequentially. Take B, for example. 
There is a dependence ``cycle'' from {\sf B[i]} to {\sf D[i]} then {\sf B[i+1]} in the subsequent 
iteration. (The cycle is shown in~\figref{fig:MPCexample}(b).) {\sf B[i]}'s cannot be computed 
in parallel because each one depends on the {\sf B[i]} value computed in the previous iteration.

Now consider the above code but slightly modified. {\sf B[i]} and {\sf C[i]} are now 
computed based on the value of {\sf D[i-2]} assigned two iterations earlier.

\end{lstlisting}
& 
~
&
\begin{lstlisting}[language=Java]
for (i=0; i<N; i++) {
  if (i%N==1) para{A[i]=f1(B[i]),...};
  if (i%2==1) para{B[i]=f2(A[i],D[i-2]),...};
  if (i%2==1) para{C[i]=f3(A[i],D[i-2]),...};
  if (i%2==1) para{D[i]=f4(B[i],C[i]),...};
}	
\end{lstlisting}

@@

{\sf B[i]} still depends on a value computed in earlier iterations, i.e., there are loop-carried dependences, 
but the value, D[i-2] is two iterations earlier. Thus, we can take parallelization blocks of 2, e.g., computing 
{\sf B[0],B[1]} in a parallel block, then {\sf B[2],B[3]}, etc.
Note however, that we cannot take a parallelization block larger than 2.
%The value of B will depend on a value being computed in parallel.

The key problem that we address in this paper is computing the maximal rate $d(s)$ allowed by program dependences 
for each statement $s$. Program dependences give rise to a partitioning of the sequence of {\sf para} blocks into a DAG of
strongly connected components. (We elaborate on this and define the partitioning precisely later in the paper.)
In the above example, there are 2 SCCs---one component consists of just the {\sf para} block that computes 
{\sf f1(B[i])} and the other component consists of the remaining three {\sf para} blocks. The three para blocks
are interdependent and hence, they are in the same component.

It is an invariant of our approach that the rates of all blocks in a SCC are the same, $d(s_1) = d(s_2) = d(\mathit{SCC})$
for each $s_1, s_2 \in \mathit{SCC}$. Unfortunately, as we argued earlier scheduling within a SCC is NP-complete, 
and also, scheduling of the components in the DAG is NP-complete. Assuming that we have an optimal schedule 
for each SCC, the sum of the costs of all SCCs is an upper bound on the optimal cost and the sum of the longest chain
in the DAG is a lower bound on the optimal cost. In the special case when the DAG is a straight line, which is the case 
is the typical case \ana{how many of the actual benchmarks???} the sum the costs of all SCCs is the \emph{minimal} cost.

\ana{Because the results are somewhat negative, i.e., thing is NP-complete, we'd need some experiments, at least 
analysis of loop structure in different benchmarks. I am pretty sure the structure is straight-line DAG with very simple SCCs, 
hence, we do have an optimal schedule anyway.}

@@

\subsubsection{MPC-Source}

MPC-Source is an SSA-based intermediate representation
of MPC programs~\cite{ccs paper}. MPC-Source is, essentially, a straight-line sequence
of ``components'', where each ``component'' is either a nested for-loop sequence (called for-loop block), 
or a simple MPC-statement. \figref{fig:example} shows an example. MPC-Source can be constructed
in a standard way from an SSA-translation of the source---each branch of an 
{\sf if}-statement is added to the sequence, the conditional becomes an MPC comparison (CMP), 
and the $\phi$-node becomes an MPC multiplexer (MUX).

MPC-Source uses standard SSA~\cite{Cytron:1991TOPLAS} which treats array writes as scalar writes, 
e.g., {\sf A[i] = f1(B[i])} is a write of a ``scalar'' variable $A$. Each {\sf A[i] = ...} and {\sf A[j] =...} will be treated 
as separate assignments to variable $A$, thus inducing {\sf A1 = ... } and {\sf A2 = ... } static single assignment
nodes. The semantics of these array assignments is different than scalar assignments in
the sense that they are ``vector assignments''; they absorb the write to the corresponding index
and copy the values at the remaining indices. However, from the point of view of analysis, including
our analysis for the purposes of scheduling and amortization, the array and scalar assignments are 
treated the same way.

The assignment to {\sf A[i]} (and to {\sf B[i], C[i]} and {\sf D[i]}) in Aiken and Nicolau's 
example will turn into {\sf A2 = f1(B1,A1)} ({\sf A1} is included in the operation because the
elements of all but index $i$ are used in the assignment) and the loop gives rise to several pseudo-$\phi$ nodes:

\begin{lstlisting}[language=Java,escapeinside=`']
// initialization of A0,B0,C0, and D0
for (i=1; i<=N; i++) {
  A1 = `$\phi$'(A0,A2);
  B1 = `$\phi$'(B0,B2);
  C1 = `$\phi$'(C0,C2);
  D1 = `$\phi$'(D0,D2);
  A2 = f1(B1,A1) // A[i]=f1(B[i]);
  B2 = f2(A1,D1,B1); // B[i]=f2(A[i],D[i-1]);
  C2 = f3(A1,D1,C1); // C[i]=f3(A[i],D[i-1]);
  D2 = f4(B2,C2,D1); // D[i]=f4(B[i],C[i]);
}	
\end{lstlisting}

The problem with standard SSA is that it looses fine-grain index information that is necessary for parallelization. 
In the above example, the analysis determines that assignment A2 depends on the previous iteration,
which entails parallelization rate of 1.

\subsubsection{Array SSA}

Therefore, we make use of Array SSA. 
Originally proposed by Knobe and Sarkar, Array SSA has been used for loop parallelization in HPC.
We adapt Array SSA for MPC-source and build our optimal loop parallelization algorithm
based on this adaptation.

Array SSA introduces a parallel array structure, denoted $@A$, associated to each array $A$. $@A$ is a ``timestamp'' 
that tracks the specific loop iteration when each $A[i]$ was last modified. We can use this information to compute a more
precise parallelization rate. 
Consider Aiken and Nicolau's example, now using Array SSA:

\begin{figure}
\begin{lstlisting}[language=Java,escapeinside=`']
for (i=1; i<=N; i++) {

@@

  A2[i] = f1(B1[i]); @A2[i] = i;
  A3 = `$\phi$'(A2,A1); @A3 = max(A2, A1);
  B2[i] = f2(A3[i],D1[i-1]); @B2[i] = i;
  B3 = `$\phi$'(B2,B1); @B3 = max(B2, B1);
  C3[i] = f3(A3[i],D1[i-1]); @C2[i] = i;
  C3 = `$\phi$'(C2,C1); @C3 = max(C2, C1);
  D3[i] = f4(B3[i],C3[i]); @D2[i] = i;
  D3 = `$\phi$'(D2,D1); @D3 = max(D2, D1); 
}
\end{lstlisting}
\caption{Aiken and Nicolau's example using Array SSA.}
\label{fig:ArraySSAAikenNicolau}
\end{figure}

There are three kinds of statements added by Array SSA. First, 
\[
{\sf A2[i] = f1(B1[i]); @A2[i] = i;}
\]
Array SSA statement {\sf @A2[i] = i;} records the iteration, $i$ when {\sf A2[i]} is written.
The index is not always $i$, it can be any $f(i)$, function of $i$, however, we assume
that the loop always iterates from 1 to $N$ with step 1. Second and third, there are
the \emph{define} $\phi$ statement  and corresponding timestamp recording statement:
\[ 
{\sf A3 = }\phi{\sf (A2,A1); @A3 = max(A2, A1);}
\]
Statement ${\sf A3 = }\phi{\sf (A2,A1);}$ is a \emph{define} $\phi$ node that merges arrays 
{\sf A2} and {\sf A1}---the semantics of the define $\phi$ is that the value at the index that is
modified is taken from ${\sf A2}$ and the values at indeces other than $i$ 
are taken from array {\sf A1}. Statement 
{\sf @A3 = max(A2, A1);} is a timestamp recording statement; it picks the latest iteration 
to assign for each index $i$---in this case, when the phi- and max- statements 
appear immediately under an array write, it assigns {\sf @A3[j] = i} if $j=i$, 
and {\sf @A3[j] = @A2[j]} otherwise.

Array SSA allows for the standard \emph{merge} $\phi$ statement at if-then-else statements. 
We may have 
\[ 
{\sf A3 = }\phi{\sf (A2,A1); @A3 = max(A2, A1);}
\]
in which case ${\sf A2}$ is the array long the true branch and ${\sf A1}$ is the array
along the false branch. 
%\ana{Add merge $\phi$}

\subsubsection{MPC-Source with Array SSA}

Merge phi nodes  e.g., ${\sf A3 = \phi(A2,A1), @A3 = max(A2, A1);}$ 
can be added to MPC-source without privacy leaks. ${\sf A3 = \phi(A2,A1)}$ is naturally translated as 
point-wise MUX for each index $i$. The timestamp statement ${\sf @A3 = max(A2, A1);}$ 
has the same meaning as in standard Array SSA---{\sf @A3[i] = max(@A2[i],@A1[i])}.
MPC-source expands and executes each arm of the if-then-else statement.
If only one arm modified the array at index $i$ {\sf @A3[i] = max(@A2[i],@A1[i]) = i}. 
Similarly, if both arms modified the array at $i$, the timestamp index again will be again $i$. 
%We note that in MPC-source the timestamp arrays {\sf @A} are known \emph{statically}, 
%unlike in standard Array SSA, where they are inherently \emph{dynamic}. 
%\ana{Double check? Make a statement that an adversary monitoring execution does not
%infer anything from SSA phi nodes.}

\subsection{Analysis Algorithm}

\subsubsection{Assumptions on MPC-Source}

To make the problem tractable we state several assumption about the loop body.
These assumptions are natural and they have been adopted in classical works as 
well~\cite{Aiken:PLDI1988}. 
These assumptions are natural and hold for all loop body sequences we have examined,
both in MPC and HPC. This is because loops that violate these assumptions would be
very difficult to understand and maintain. Moreover, if loops use arrays in readonly way, 
which is the case in the vast majority of cases particularly in MPC, the assumptions automatically 
hold true.
~\ana{This statement is true for all loops I have seen but we do need 
a more systematic way of studying loops and reporting on structure.}

%Array writes.~\ana{We have to see what the distinction was, regular vs. irregular loops}:
\begin{enumerate}
\item For each array read of $A$: $... = g(A[f(i)])$ and corresponding write to $A$: $A[f'(i)] = ...$
we require that $f(i) = f'(i)$ for every $0 \le i < N$. Here $i$ is the induction variable and $N$ is
the loop bound.

\item Let $A[f_1(i)]...A[f_k(i)]$ be a sequence of writes to $A$ in the loop body sequence. 
%\ana{How is the sequence determined?} with $A[f_k(i)]$ being the most recent one. 
As usual, here $i$ denotes the induction variable. We require that for every $0 \le i < N, k,l$, $f_k(i) \neq f_l(i)$, 
or in other words, those writes access different elements of the array. Furthermore, we require 
that $f_k(i) \neq f_l(j)$, or in other words, writes across different iterations access different array 
elements. %\ana{Needs rewording. Can be simplified.}

\item Our last assumption is the ``constant rate'' assumption. It states that a statement 
$s$ in the loop body sequence depends on itself at a ``constant rate'', i.e., there cannot be a 
case where $s(i)$ depends on $s(j), j<i$ and on $s(k), k<i$ s.t. $i \neq j$. (Here $s(i)$
denotes the execution of $s$ in iteration $i$.)
More formally, let $s(j) \dashrightarrow \cdot \longrightarrow s(i)$ be a dependence chain,
where the dashed arrow denotes the loop carried dependence. For every 
$s(j) \dashrightarrow \cdot \longrightarrow s(i)$ we have $i = j+m$ for some constant $m$.

%Constant loop schedule. Let $S$ be a loop schedule, i.e., a directed acyclic graph that defines "executes before"
%scheduling constraints among the statements of the loop body. We require that the 
%parallelization of the loop schedule obeys the constant loop schedule constraint: each instance of $S$
%in the loop linearization obeys the scheduling constraints defined by the DAG of $S$. \ana{Justify why!}

\subsubsection{Index Invariant}

The \emph{index invariant} makes an assertion about when {\sf A[i]} is written as a function of $i$.
It is essential in computing the rate of parallelization $d(s)$. \figref{fig:ArraySSAMPCAikenNicolau}
shows the index invariant on ${\sf @A1}$ (all other arrays have the exact same index invariant). 
The index invariant can be specified by the programmer, as it is a form of a loop invariant or it can 
be inferred. \ana{Can we infer the invariant???}

%\ana{Can we compute this index easily: A[f(j)] = i, if j>=i, A[f(j)] = 0 otherwise?
%Under what assumptions this is true???}. \ana{We will use the Array SSA stuff to compute index; 
%can get rid of it once we.}

\subsubsection{Computing $d(s)$}

We assume a loop body that consists of three-address MPC statements:
\[ v1 = v2 \oplus v3 \]
MUX statements:
\[ v1 = MUX(v2, v3, v4) \]
and $\phi$-nodes:
\[ v1 = \phi(v2, v3) \]
where $v1,v2$, $v3$, and $v4$ are either scalar variables, e.g., {\sf x} or array elements {\sf A[f(i)]}.
Define phi- and max- statements are elided from the representation, as
their only purpose is to compute the invariant. Merge phi-statements are translated
into point-wise MUX statements, as is standard in MPC.

\ana{Pseudo phi nodes!}

\begin{figure}
\begin{lstlisting}[language=Java,escapeinside=`']
for (i=1; i<=N; i++) {

  // Index Invariant: @A1[j] == j if `$j < i$', @A1[j] == 0 otherwise

  A1 = `$\phi$'(A3,A0); 
  B1 = `$\phi$'(B3,B0); 
  C1 = `$\phi$'(C3,C0);
  D1 = `$\phi$'(D3,D0);

  A2[i] = f1(B1[i]); 
  B2[i] = f2(A3[i],D1[i-1]); 
  C3[i] = f3(A3[i],D1[i-1]); 
  D3[i] = f4(B3[i],C3[i]); 
}
\end{lstlisting}
\caption{Aiken and Nicolau's example using Array SSA.}
\label{fig:ArraySSAMPCAikenNicolau}
\end{figure}

%\ana{Figure out how to reason about cycles based on Array SSA.}

Recall that the loop body is a linear sequence of statements. To compute $d(s)$ of statement $s$, 
we follow def-use chains backwards collecting all reachable pseudo-phi node. If there is a psudo-phi 
node of a scalar, then $d(s) = 1$; for an array we retrieve the value $A[f(i)]$ from the invariant. \ana{formalize, polish!}

\ana{Define Schedule $S$.}

\ana{Add Algorithm}

\ana{Prove main theorem, that this schedule is of minimal cost. I.e., under the restriction of "constant loop body schedule", there
does not exist a schedule $S'$}

for a statement x = f(y,z), follow the reverse def-use link.
\end{comment}

\subsection{High-level CFG to MPC $G$}
\label{CFG}

\subsubsection{From $\phi$-nodes to MUX-nodes}

\ana{Difficulties: Translate to MUX nodes, adapt classical SSA algorithm, arrays, reads and writes!}

\begin{figure}
\small
\[
\begin{array}{l@{~~~~~}l}
  \begin{array}{l@{~}l@{~~~}l}
  s & ::= s ; s \\ 
  & \mid \x = \y \; \mid \x=\y \; \op \; \z \mid {\sf A[f(i)] = x} \; \mid {\sf x = A[f(i)]} & \mathit{assignment} \\
%  & \mid \x = \phi(\y,\z) \\
  & \mid {\sf for} \; (i = 0; i \le \mathit{Const}; \mathit{i\!+\!+}) \; \{ \; s \; \} \\
  & \mid {\sf if} \; (\x \; \cop \; \y) \; \{ \; s \; \} \; {\sf else} \; \{ \; s \; \} \\
 % \bool &::= \x \; \cop \; \y & \mathit{comparison} \\
  \op &::= + \mid - \mid * \mid / & \mathit{arithmetic} \; \mathit{operator} \\
  \cop &::=  \; == \; \mid \; != \; \mid \; < \; \mid \; \le & \mathit{comparison} \; \mathit{operator}

  \end{array}
\end{array}
\]
%\vspace{-0.2in}
\caption{IMP syntax as defined in \cite{Ishaq2019}: $s$ is a sequence of statements. 
\code{x}, \code{y}, \code{z}, and $i$, are variables.
} 
\label{fig:syntax}
\end{figure}

We consider a standard IMP-like imperative language. The syntax is given in~\figref{fig:syntax}. 
Without loss of generality, we assume that the syntax gives rise to a CFG and we can apply the 
Cytron et al.'s Algorithm for the efficient computation of SSA over this CFG~\cite{Cytron:1998}.

In particular, we are interested in a sequence of statements that form a loop body, i.e., while 
the outermost statement is a {\sf for} the remaining statements are assignments and {\sf if}s. 
There are two principal hurdles. One is to translate the CFG into SSA form, where $\phi$ nodes 
are turned into MUX nodes. One cannot simply reuse the classical algorithm by Cytron et. al. 
because $V_3 = \phi(V_1,V_2)$ nodes do not retain information about the predicate that triggers 
$V_1$ vs. $V_2$; all $\phi$ nodes represent is conditional control flow, i.e., that $V_3$ is either 
$V_1$ or $V_2$. Another issue is that SSA $\phi$ nodes can combine multiple conditionals. 
To illustrate, consider the code

\begin{minted}[fontsize=\footnotesize, linenos, numbersep=5pt, escapeinside=||]{java}
z = 0;
...
if (x > 0) {
  if (y > 0) {
    z = 1;
  }  else {
    z = -1;
  }
}
print z
\end{minted}
Line 10 is the dominance frontier of the assignments to \z, and Cytron et. al.'s algorithm
creates a $\phi$ node at 10: $z_3 = \phi(z_0, z_1, z_2)$. There is no information retained
that $z_0$ is triggered when $x \le 0$, $z_1$ is triggered when $x > 0 \; \wedge \; y > 0$, 
and $z_2$ is triggered when $x > 0 \; \wedge \; y \le 0$. The Multiplexer at the join point
needs this information however in order to assign the correct value to $z_3$ at runtime.

We propose a simple extension to the classical algorithm. Recall from Cytron et. al.'s 
that efficient static single assignment computation has two phases, phase one is the 
placement of phi-nodes at dominance frontiers of assignments, and phase two is the 
renaming of variables $V$ into $V_1$, $V_2$, etc.

\ana{Here we need an explanation for lines 23-45. Turning n-ary phi-nodes into binary MUX nodes, 
just the same classical algorithm with minor extension.}

\ana{Example of CFGs for above example.}

Once we have the graph in SSA form with \ana{MORE}

\begin{algorithm}
\caption{Determine Graph edges}
\begin{algorithmic}[1]
%\REQUIRE $c \neq null$
%\STATE $L \leftarrow get\_nodes(c)$
%\FORALL{$n$ in $L$}
%\STATE $...$
%\ENDFOR
\FORALL {variables $V$}
\STATE $C(V) \leftarrow 0$
\STATE $S(V) \leftarrow \mathit{EmptyStack}$
\STATE $P \leftarrow \mathit{EmptyStack}$
\ENDFOR
\STATE call SEARCH($\mathit{Entry}$) \\
\STATE ... \\
SEARCH(X) : \\
\FORALL {statements $A$ in $X$} 
\IF {$A$ is a $\phi$-assignment}
   \STATE {remove $A$ and continue}
\ENDIF
\FORALL {variables $V$ in $\mathit{RHS}(A)$}
   \STATE {replace $V$ with $V_i$ where $i = \mathit{Top}(S(V))$}
\ENDFOR

\FORALL {variable $V$ in $\mathit{LHS}(A)$}
   \STATE { $i \gets C(V)$} 
   \STATE {replace $V$ with $V_i$ in $\mathit{LHS}(A)$}
   \STATE {push $i$ onto $S(V)$}
   \STATE {$C(V) \gets i+1$}
\ENDFOR
\ENDFOR

\FORALL {$Y \in Succ(X)$}
\IF {$X$ is a conditional node}
   \STATE {push $\mathit{Cond}, \mathit{Branch}$ onto $P$ // $\mathit{Branch}$ is $T$ if $X \rightarrow Y$ is the True branch, $F$ otherwise}     
\ENDIF   
     \STATE{// $Y$ is not done if $Y$ has an incomplete MUX predecessor or an original CFG predecessor \ana{Better?}}
      \IF {$Y$ has $\phi$-assignments and $Y$ is not done} 
          \STATE {$\mathit{Cond}, \mathit{Branch} \gets$ pop $P$}
          \FORALL {$\phi$-assignments in $Y$}
               \IF {$Z\!: V = \mathit{MUX}(Cond, ..., ...) \in \mathit{Pred}(Y)$}
                  %\STATE {replace $V$ in $RHS(Z)$ with $V_i$, where $i = Top(S(V)$}
                  \STATE {remove edge $X\rightarrow Y$, add edge $X\rightarrow Z$ to $\mathit{CFG}$}
                  \STATE {call SEARCH($Z$) // updates $V$'s in LHS and RHS} 
               \ELSE
                   \STATE {$i = Top(S(V))$}
                   \IF {$\mathit{Branch} == T$}
                      \STATE {create new node $Z\!: V = \mathit{MUX}(\mathit{Cond}, V_i, V)$}
                   \ELSE 
                      \STATE {create new node $Z\!: V = \mathit{MUX}(\mathit{Cond}, V, V_i)$}
                   \ENDIF   
                   \STATE {remove edge $X\rightarrow Y$, add edge $X\rightarrow Z$ to $\mathit{CFG}$} 
               \ENDIF
          \ENDFOR 
      \ENDIF
\ENDFOR
\FORALL {$Y \in \mathit{Children}(X)$}
   \STATE {call SEARCH($Y$)}
\ENDFOR
\FORALL {assignments $A$ in $X$}
   \FORALL {variables $V$ in oldLHS($A$)}
       \STATE {pop S(V)}
   \ENDFOR
\ENDFOR
\end{algorithmic}
\label{alg:algorithm1}
\end{algorithm}

\subsubsection{Handling of Arrays}
\label{???}

So far we focused on the classical SSA analysis and the extension to create MUX nodes. 
Classical SSA treats arrays as scalars, i.e., an array write, e.g., {\sf A[i] = x+y} is considered
a modification of {\sf A} and warrants its own indexed variable ${\sf A}_k$. A later use of {\sf A}, 
even one such as {\sf x = A[i+1]} where there is no def-use link from {\sf A[i] = x+y}, will be 
considered a use of ${\sf A}_k$. This hinders parallelization. To illustrate, consider our running
example in~\figref{fig:MPCexample}:
\begin{minted}[fontsize=\footnotesize, numbersep=5pt, escapeinside=||]{java}
for (...) {
  A: A[i] = f_1(B[i]);
  B: B[i] = f_2(A[i]);
  ...
}
\end{minted}
Since {\sf A} and {\sf B} are treated as standard scalars, the dependence analysis will record
a dependence from {\sf B} to {\sf A} via the back edge. As a result, statement {\sf A} wouldn't 
be vectorized because each (i+1)-th iteration depends on {\sf B} in the i-th iteration,
which depends on {\sf A} in the i-th iteration.

To mitigate against this, we propose a technique, similar to Array-SSA though much simpler and 
less general, which breaks certain dependence edges and allows for better vectorization. 
\ana{Have to make a claim how 
often this technique resolves dependences perfectly, and how often it defaults to standard SSA.}

\ana{TODO: Summary of the technique}

\paragraph {Equivalence and Non-equivalence} Let $f(i)$ and $f'(i)$ be two index functions on induction variables $i$, 
e.g., we have array accesses ${\A}[f(i)]$ and ${\A}[f'(i)]$. We say that $f(i) \cong f'(i)$ iff for every $1 \le i \le N$,
$f(i) = f'(i)$. We say that $f(i) \ncong f'(i)$ iff for every $1 \le i \le N$, $f(i) \neq f'(i)$.

\paragraph{Array Well-formedness Conditions} Let ${\A}$ be an array mentioned in the loop body. 
We will treat array accesses ${\A}[f(i)]$ as distinct scalar variables if we can show that the loop body is
well-formed with respect to \A. Otherwise, we treat array {\A} as a single variable, thus defaulting to the 
case of standard SSA. The well-formedness conditions are as follows: \ana{This needs some work!}

\begin{enumerate}
\item For each pair of array writes to \A, $n_1\!: A[f(i)] = ... $ and $n_2\!: A[f'(i)] = ... $, we have either $f(i) \cong f'(i)$ or 
$f(i) \ncong f'(i)$.

\item For each pair $n_1\!: A[f(i)] = ... $ and $n_2\!: ... = A[f'(i)]$ such that $n_1$ reaches $n_2$ on a forward path, 
we have either $f(i) \cong f'(i)$ or $f(i) \ncong f'(i)$.

%\item For each pair $n_1\!: A[f(i)] = ... $ and $n_2\!: A[f'(i)] = ... $ either $f(i-d) \cong f'(i)$ for some unique $d > 0$, 
%or $f(i-d) \ncong f'(i)$ for every $d > 1$.
